---
title: "Causal structures: confounder, mediator, collider, M-bias"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal structures: confounder, mediator, collider, M-bias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  message   = FALSE
)
library(kagu)
set.seed(42)
N <- 500
```

This vignette compares naive OLS against Kagu's do-calculus estimates on four
canonical graph structures where conditioning on the wrong set of variables
leads to bias — and a fifth where *nothing* works, to show the method's limits.

**How to read each section.** For every structure we print two ordinary
least-squares estimates of the `x → y` coefficient — one *naive* (`y ~ x`) and
one *adjusted* (`y ~ x + ...`) — then Kagu's interventional estimate as a tidy
table. In that table `mean` is the posterior mean effect and
`hdi_lower`/`hdi_upper` bound the 90% highest density interval (the range
containing 90% of the posterior probability — the credible values). The lesson
each time: OLS is right only with the *correct* adjustment set, which you can
only know from the causal structure; Kagu reads that structure off the DAG and
applies the do-operator directly.

---

## 1 — Confounder

```
z → x → y
z → y
```

The confounder `z` causes both `x` and `y`. Naive OLS of `y ~ x` without
adjusting for `z` is biased; Kagu uses the full structural model.

True effect of `x` on `y`: **+1.5**.

```{r confounder-data}
z <- rnorm(N)
x <- 1.0 * z + rnorm(N, sd = 0.5)
y <- 1.5 * x + 1.0 * z + rnorm(N, sd = 0.5)
df_confounder <- data.frame(z = z, x = x, y = y)
```

```{r confounder-ols}
cat("OLS (naive):   ", coef(lm(y ~ x, df_confounder))[["x"]], "\n")
cat("OLS (adjusted):", coef(lm(y ~ x + z, df_confounder))[["x"]], "\n")
```

```{r confounder-kagu, results = "hide"}
model_confounder <- KaguModel$new(dag = list(z = c(), x = c("z"), y = c("x", "z")))
model_confounder$fit(df_confounder)
```

```{r confounder-dag, fig.height = 3}
model_confounder$plot_dag(node_pos = list(
  z = c(0, 0), x = c(1, -1), y = c(1, 1)
))
```

```{r confounder-effect}
model_confounder$effects("x", "y", values = c(0, 1))$summary()
```

Notice the naive OLS estimate sits well above 1.5 — the confounding path
`x ← z → y` leaks association into the slope. The *adjusted* OLS is correct here
only because we happened to condition on `z`. Kagu's table recovers ≈ 1.5 with a
90% HDI comfortably around the truth, because the structural model represents
`z` as a common cause and the do-operator severs the incoming edge to `x`.

---

## 2 — Mediator

```
x → m → y
```

The mediator `m` lies on the causal path. Naive OLS of `y ~ x + m` blocks
the path and gives a biased (near-zero) estimate of the total effect.

True total effect of `x` on `y`: **−1.6** (= 0.8 × −2).

```{r mediator-data}
x <- rnorm(N)
m <- 0.8 * x + rnorm(N, sd = 0.5)
y <- -2.0 * m + rnorm(N, sd = 0.5)
df_mediator <- data.frame(x = x, m = m, y = y)
```

```{r mediator-ols}
cat("OLS (naive):   ", coef(lm(y ~ x, df_mediator))[["x"]], "\n")
cat("OLS (adjusted):", coef(lm(y ~ x + m, df_mediator))[["x"]], "\n")
```

```{r mediator-kagu, results = "hide"}
model_mediator <- KaguModel$new(dag = list(x = c(), m = c("x"), y = c("m")))
model_mediator$fit(df_mediator)
```

```{r mediator-dag, fig.height = 2.2}
model_mediator$plot_dag(node_pos = list(
  x = c(0, 0), m = c(0, 1), y = c(0, 2)
))
```

```{r mediator-effect}
model_mediator$effects("x", "y", values = c(0, 1))$summary()
```

Here the roles flip: *naive* OLS is correct and *adjusted* OLS is biased. Adding
`m` to the regression conditions on the mediator and blocks the very path the
effect travels along, collapsing the estimate toward zero. Kagu propagates
through `m` automatically, so its table recovers the total effect ≈ −1.6 — you
never have to reason about which covariates to include.

---

## 3 — Collider

```
x → c ← y
```

The collider `c` has two causes. `x` and `y` are independent — the true
effect of `x` on `y` is **0**. Conditioning on `c` opens a spurious path and
induces bias.

```{r collider-data}
x <- rnorm(N)
y <- rnorm(N)
c <- 0.7 * x + 0.7 * y + rnorm(N, sd = 0.5)
df_collider <- data.frame(x = x, y = y, c = c)
```

```{r collider-ols}
cat("OLS (naive):   ", coef(lm(y ~ x, df_collider))[["x"]], "\n")
cat("OLS (adjusted):", coef(lm(y ~ x + c, df_collider))[["x"]], "\n")
```

```{r collider-kagu, results = "hide"}
model_collider <- KaguModel$new(dag = list(x = c(), y = c(), c = c("x", "y")))
model_collider$fit(df_collider)
```

```{r collider-dag, fig.height = 2.2}
model_collider$plot_dag(node_pos = list(
  x = c(0, 0), c = c(0, 1), y = c(0, 2)
))
```

```{r collider-effect}
model_collider$effects("x", "y")$summary()
```

Naive OLS is correct (≈ 0), but adjusting for the collider `c` induces a
spurious association — conditioning on a common effect makes its causes appear
related. Kagu returns an effect whose HDI straddles zero, because `x` has no
directed path to `y` in the DAG and the propagation simply finds nothing to
carry.

---

## 4 — M-bias

```
a → x    b → y
a → m ← b
```

`m` is a collider on a non-causal path. The true effect of `x` on `y` is
**+1.0**. Conditioning on `m` opens a spurious path via `a` and `b`,
biasing a naive adjusted OLS estimate.

```{r mbias-data}
a <- rnorm(N)
b <- rnorm(N)
x <- 1.0 * a + rnorm(N, sd = 0.5)
y <- 1.0 * x + 1.0 * b + rnorm(N, sd = 0.5)
m <- 0.7 * a + 0.7 * b + rnorm(N, sd = 0.5)
df_mbias <- data.frame(a = a, b = b, x = x, y = y, m = m)
```

```{r mbias-ols}
cat("OLS (naive):   ", coef(lm(y ~ x, df_mbias))[["x"]], "\n")
cat("OLS (adjusted):", coef(lm(y ~ x + m, df_mbias))[["x"]], "\n")
```

```{r mbias-kagu, results = "hide"}
model_mbias <- KaguModel$new(dag = list(
  a = c(), b = c(),
  x = c("a"), y = c("x", "b"),
  m = c("a", "b")
))
model_mbias$fit(df_mbias)
```

```{r mbias-dag, fig.height = 4}
model_mbias$plot_dag(node_pos = list(
  a = c(0, -2), b = c(0, 2),
  m = c(1,  0),
  x = c(2, -2), y = c(2, 2)
))
```

```{r mbias-effect}
model_mbias$effects("x", "y", values = c(0, 1))$summary()
```

M-bias is the subtle one: `m` is a *pre-treatment* collider, so the usual
instinct to "adjust for everything measured beforehand" backfires. Adjusted OLS
opens the path `x ← a → m ← b → y` and biases the estimate, while naive OLS
(≈ 1.0) is fine. Kagu, working from the correct structural model, never adjusts
for `m` during propagation and its table recovers ≈ 1.0.

---

## 5 — Latent confounder (where everything fails)

```
u → x → y
u → y       (u is unobserved)
```

So far the structural model has always won. This example shows where it
*cannot*, and why that is the honest answer rather than a shortcoming. The
common cause `u` is now **latent** — it drives both `x` and `y` but is never
recorded. The true direct effect of `x` on `y` is **+1.5**.

The structure we *wish* we could fit — `u` is the unobserved common cause:

```{r latent-dag, fig.height = 3}
kagu_plot_dag(
  list(u = c(), x = c("u"), y = c("x", "u")),
  node_pos = list(u = c(0, 0), x = c(1, -1), y = c(1, 1))
)
```

```{r latent-data}
u <- rnorm(N)                          # the hidden confounder
x <- 1.0 * u + rnorm(N, sd = 0.5)
y <- 1.5 * x + 1.0 * u + rnorm(N, sd = 0.5)
df_latent <- data.frame(x = x, y = y)  # u is NOT included
```

With only `x` and `y` recorded, the only regression we can run is the naive one
— there is no `u` column to adjust for:

```{r latent-ols}
cat("OLS (naive):   ", coef(lm(y ~ x, df_latent))[["x"]], "\n")
cat("OLS (adjusted): unavailable — 'u' is not measured\n")
```

The naive slope is biased upward, exactly as in the confounder example — but
this time there is no adjustment available to repair it.

Can the structural model rescue us? Only if we can supply the confounder.
Writing the *correct* DAG and asking Kagu to fit it fails fast, by design:

```{r latent-kagu-correct, error = TRUE}
model_latent <- KaguModel$new(dag = list(u = c(), x = c("u"), y = c("x", "u")))
model_latent$fit(df_latent)
```

Kagu refuses to invent data for an unmeasured node, and the message spells out
the consequence. The alternative is to drop `u` from the DAG so the model fits —
but then it is structurally identical to the naive regression, and just as
biased:

```{r latent-kagu-naive, results = "hide"}
model_latent_naive <- KaguModel$new(dag = list(x = c(), y = c("x")))
model_latent_naive$fit(df_latent)
```

```{r latent-kagu-naive-effect}
model_latent_naive$effects("x", "y", values = c(0, 1))$summary()
```

The estimate lands near the inflated OLS value, not the true 1.5, and the HDI
does not cover it. This is the key lesson: **unobserved confounding cannot be
undone from the data alone.** Neither regression nor a graphical model recovers
the truth when a common cause is missing — the effect is simply *not
identifiable* from these variables. The remedies all live outside the dataset:
measure the confounder, find a valid instrument, or exploit a front-door path.

---

## Summary

| Structure         | True effect | OLS (naive) | OLS (adjusted) | Kagu                   |
|-------------------|-------------|-------------|----------------|------------------------|
| Confounder        | +1.5        | Biased ↑    | Correct        | ✓ Correct              |
| Mediator          | −1.6        | Correct     | Biased ≈ 0     | ✓ Correct              |
| Collider          | 0.0         | Correct     | Biased ≠ 0     | ✓ Correct              |
| M-bias            | +1.0        | Correct     | Biased         | ✓ Correct              |
| Latent confounder | +1.5        | Biased ↑    | Unavailable    | ⚠ Stops and warns      |

OLS is correct only when it happens to use the right adjustment set. Kagu
encodes the causal structure explicitly and applies the do-operator — correct
by design *whenever the effect is identifiable*.

**Read the last row as a safety feature, not a shortcoming.** No method —
regression or graphical — can recover an effect that the data simply cannot
identify; that is a property of the problem, not of the tool. The difference is
in *how each one behaves at that boundary*. OLS hands back a confident,
biased number with no hint that anything is wrong. Kagu instead detects the
unmeasured node, **stops before fitting, and emits a clear warning** explaining
why the estimate would be untrustworthy and what to do about it (measure the
confounder, find an instrument, or use a front-door path). A loud, actionable
error is exactly what you want here — it turns a silent statistical trap into a
visible, fixable one.
