---
title: "Causal structures: confounder, mediator, collider, M-bias"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal structures: confounder, mediator, collider, M-bias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  message   = FALSE
)
library(kagu)
set.seed(42)
N <- 500
```

This vignette compares naive OLS against Kagu's do-calculus estimates on four
canonical graph structures where conditioning on the wrong set of variables
leads to bias.

---

## 1 — Confounder

```
z → x → y
z → y
```

The confounder `z` causes both `x` and `y`. Naive OLS of `y ~ x` without
adjusting for `z` is biased; Kagu uses the full structural model.

True effect of `x` on `y`: **+1.5**.

```{r confounder-data}
z <- rnorm(N)
x <- 1.0 * z + rnorm(N, sd = 0.5)
y <- 1.5 * x + 1.0 * z + rnorm(N, sd = 0.5)
df_confounder <- data.frame(z = z, x = x, y = y)
```

```{r confounder-ols}
cat("OLS (naive):   ", coef(lm(y ~ x, df_confounder))[["x"]], "\n")
cat("OLS (adjusted):", coef(lm(y ~ x + z, df_confounder))[["x"]], "\n")
```

```{r confounder-kagu, results = "hide"}
model_confounder <- KaguModel$new(dag = list(z = c(), x = c("z"), y = c("x", "z")))
model_confounder$fit(df_confounder)
```

```{r confounder-dag}
model_confounder$plot_dag()
```

```{r confounder-effect}
model_confounder$effects("x", "y", values = c(0, 1))$summary()
```

Kagu recovers the true effect because the structural model explicitly
represents `z` as a common cause.

---

## 2 — Mediator

```
x → m → y
```

The mediator `m` lies on the causal path. Naive OLS of `y ~ x + m` blocks
the path and gives a biased (near-zero) estimate of the total effect.

True total effect of `x` on `y`: **−1.6** (= 0.8 × −2).

```{r mediator-data}
x <- rnorm(N)
m <- 0.8 * x + rnorm(N, sd = 0.5)
y <- -2.0 * m + rnorm(N, sd = 0.5)
df_mediator <- data.frame(x = x, m = m, y = y)
```

```{r mediator-ols}
cat("OLS (naive):   ", coef(lm(y ~ x, df_mediator))[["x"]], "\n")
cat("OLS (adjusted):", coef(lm(y ~ x + m, df_mediator))[["x"]], "\n")
```

```{r mediator-kagu, results = "hide"}
model_mediator <- KaguModel$new(dag = list(x = c(), m = c("x"), y = c("m")))
model_mediator$fit(df_mediator)
```

```{r mediator-dag}
model_mediator$plot_dag()
```

```{r mediator-effect}
model_mediator$effects("x", "y", values = c(0, 1))$summary()
```

Kagu propagates through `m` automatically and recovers the total causal effect.

---

## 3 — Collider

```
x → c ← y
```

The collider `c` has two causes. `x` and `y` are independent — the true
effect of `x` on `y` is **0**. Conditioning on `c` opens a spurious path and
induces bias.

```{r collider-data}
x <- rnorm(N)
y <- rnorm(N)
c <- 0.7 * x + 0.7 * y + rnorm(N, sd = 0.5)
df_collider <- data.frame(x = x, y = y, c = c)
```

```{r collider-ols}
cat("OLS (naive):   ", coef(lm(y ~ x, df_collider))[["x"]], "\n")
cat("OLS (adjusted):", coef(lm(y ~ x + c, df_collider))[["x"]], "\n")
```

```{r collider-kagu, results = "hide"}
model_collider <- KaguModel$new(dag = list(x = c(), y = c(), c = c("x", "y")))
model_collider$fit(df_collider)
```

```{r collider-dag}
model_collider$plot_dag()
```

```{r collider-effect}
model_collider$effects("x", "y")$summary()
```

Kagu returns an effect near zero — `x` has no causal path to `y` in this DAG.

---

## 4 — M-bias

```
a → x    b → y
a → m ← b
```

`m` is a collider on a non-causal path. The true effect of `x` on `y` is
**+1.0**. Conditioning on `m` opens a spurious path via `a` and `b`,
biasing a naive adjusted OLS estimate.

```{r mbias-data}
a <- rnorm(N)
b <- rnorm(N)
x <- 1.0 * a + rnorm(N, sd = 0.5)
y <- 1.0 * x + 1.0 * b + rnorm(N, sd = 0.5)
m <- 0.7 * a + 0.7 * b + rnorm(N, sd = 0.5)
df_mbias <- data.frame(a = a, b = b, x = x, y = y, m = m)
```

```{r mbias-ols}
cat("OLS (naive):   ", coef(lm(y ~ x, df_mbias))[["x"]], "\n")
cat("OLS (adjusted):", coef(lm(y ~ x + m, df_mbias))[["x"]], "\n")
```

```{r mbias-kagu, results = "hide"}
model_mbias <- KaguModel$new(dag = list(
  a = c(), b = c(),
  x = c("a"), y = c("x", "b"),
  m = c("a", "b")
))
model_mbias$fit(df_mbias)
```

```{r mbias-dag}
model_mbias$plot_dag()
```

```{r mbias-effect}
model_mbias$effects("x", "y", values = c(0, 1))$summary()
```

Kagu, working from the correct structural model, never adjusts for `m` during
effect propagation and recovers the true effect.

---

## Summary

| Structure  | True effect | OLS (naive) | OLS (adjusted) | Kagu |
|------------|-------------|-------------|----------------|------|
| Confounder | +1.5        | Biased ↑    | Correct        | ✓    |
| Mediator   | −1.6        | Correct     | Biased ≈ 0     | ✓    |
| Collider   | 0.0         | Correct     | Biased ≠ 0     | ✓    |
| M-bias     | +1.0        | Correct     | Biased         | ✓    |

OLS is correct only when it happens to use the right adjustment set. Kagu
encodes the causal structure explicitly and applies the do-operator — correct
by design.
