---
title: "Causal structure discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal structure discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  message   = FALSE
)
library(kagu)
set.seed(42)
```

The other vignettes take the DAG as given. But the DAG is itself a *hypothesis*
about how a system works, and often it is exactly what we want to learn. Kagu
can treat structure discovery as Bayesian inference over graphs: fit a GCM to
every candidate DAG, and turn the marginal likelihoods into a posterior
distribution over structures,

$$
P(G \mid X) \;\propto\; P(X \mid G)\, P(G).
$$

Here $P(X \mid G)$ is the marginal likelihood of the data under graph $G$
(computed in closed form from the per-node Gaussian-process fits) and $P(G)$ is a prior over
structures (currently uniform). Rather than returning a single "best" graph,
this returns a full probability distribution — an honest representation of what
the data can and cannot tell us about causal structure.

## A worked example

We simulate a four-variable system with a known structure. `education` is a root
cause of both `skill` and `network`; `income` is driven by `skill` and
`network`. There is **no** direct `education → income` edge, and `skill` and
`network` are not directly linked.

```{r truth, fig.height = 3.5}
true_dag <- list(
  education = c(),
  skill     = "education",
  network   = "education",
  income    = c("skill", "network")
)

kagu_plot_dag(true_dag, node_pos = list(
  education = c(0,  0),
  skill     = c(1, -1),
  network   = c(1,  1),
  income    = c(2,  0)
))
```

```{r simulate}
n         <- 400
education <- rnorm(n)
skill     <- 0.8 * education + rnorm(n, sd = 0.5)
network   <- 0.6 * education + rnorm(n, sd = 0.5)
income    <- 1.0 * skill + 0.7 * network + rnorm(n, sd = 0.5)

df <- data.frame(education, skill, network, income)
```

## Running the search

The number of DAGs grows explosively with the number of variables, so we narrow
the space with background knowledge. Here we know a **temporal ordering** —
education comes before skill and professional network, which come before income
— and encode it through `disallowed`, forbidding any edge that would point
backwards in time. This is exactly the kind of domain knowledge a researcher
brings to a problem, and it makes the search tractable.

```{r discover}
disallowed <- list(
  c("skill",   "education"), 
  c("network", "education"), 
  c("income", "education")
)

result <- KaguModel$discover(df, disallowed = disallowed)
```

```{r print-result}
result
```

The search evaluated every DAG consistent with the ordering, but — thanks to the
Markov factorisation — only had to fit each unique `node ~ parents` model once
(far fewer fits than DAGs).

## The posterior over DAGs

`$summary()` ranks the structures by posterior probability. Each row is a whole
DAG; `posterior_prob` is $P(G \mid X)$.

```{r summary}
result$summary()
```

The histogram makes the shape of the posterior clear. With the true
data-generating DAG highlighted, we can see how much of the probability mass it
captures.

```{r plot}
result$plot(true_dag = true_dag)
```

## Edge probabilities

Often we care less about the single best graph than about *specific* causal
questions — does this edge exist, and in which direction? `$edge_probabilities()`
marginalises over the whole posterior to give the probability of each directed
edge, $P(\text{from} \to \text{to} \mid X) = \sum_{G \ni \text{edge}} P(G \mid X)$.

```{r edges}
result$edge_probabilities()
```

The four true edges (`education→skill`, `education→network`, `skill→income`,
`network→income`) should carry almost all the probability, while spurious edges
(a direct `education→income`, or `skill→network`) carry very little.

## Effects under structural uncertainty

When the structure is uncertain, committing to a single DAG — even the most
probable one — throws away that uncertainty. Instead, query the effect directly
on the `DiscoveryResult`: it averages over the *entire* posterior of DAGs,
weighting each by $P(G \mid X)$,
$$
P(Q \mid X) = \sum_G P(Q \mid X, G)\, P(G \mid X).
$$
The call signature and the returned object are exactly those of the single-DAG
`model$effects()`, so everything downstream — summaries, plots — works the same.

```{r effect}
result$effects("education", "income")$summary()
```

This estimate carries both the parameter uncertainty within each model and the
structural uncertainty between models. DAGs in which `education` has no causal
path to `income` simply contribute a zero effect, exactly as they should.

## A note on Markov equivalence

Because we supplied a temporal ordering, every candidate DAG shares the same
variable order, which resolves the ambiguity between statistically equivalent
structures. Without such constraints, *Markov-equivalent* DAGs — those that
imply the same conditional independences, such as `x → y → z`, `x ← y ← z`, and
`x ← y → z` — are indistinguishable from observational data alone, and the
posterior would (correctly) spread its mass evenly across them. That is not a
limitation of the method but an honest reflection of what passive observation
can determine. Distinguishing such structures requires either prior knowledge
(as here) or interventional data.

The uniform prior used here can, in future, be replaced by priors that favour
sparsity or encode softer structural beliefs — the posterior machinery is
unchanged.
