---
title: "Causal structure discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal structure discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  message   = FALSE
)
library(kagu)
set.seed(42)
```

The other vignettes take the DAG as given. But the DAG is itself a *hypothesis*
about how a system works, and often it is exactly what we want to learn. Kagu
can treat structure discovery as Bayesian inference over graphs: fit a GCM to
every candidate DAG, and turn the marginal likelihoods into a posterior
distribution over structures,

$$
P(G \mid X) \;\propto\; P(X \mid G)\, P(G).
$$

Here $P(X \mid G)$ is the marginal likelihood of the data under graph $G$
(computed in closed form from the per-node Gaussian-process fits) and $P(G)$ is a prior over
structures (currently uniform). Rather than returning a single "best" graph,
this returns a full probability distribution — an honest representation of what
the data can and cannot tell us about causal structure.

## A worked example

We simulate a four-variable system with a known structure. `education` is a root
cause of both `skill` and `network`; `income` is driven by `skill` and
`network`. There is **no** direct `education → income` edge, and `skill` and
`network` are not directly linked.

```{r truth, fig.height = 3.5}
true_dag <- list(
  education = c(),
  skill     = "education",
  network   = "education",
  income    = c("skill", "network")
)

kagu_plot_dag(true_dag, node_pos = list(
  education = c(0,  0),
  skill     = c(1, -1),
  network   = c(1,  1),
  income    = c(2,  0)
))
```

```{r simulate}
n         <- 200
education <- rnorm(n)
skill     <- 0.8 * education + rnorm(n, sd = 0.6)
network   <- 0.6 * education + rnorm(n, sd = 0.6)
income    <- 1.0 * skill + 0.7 * network + rnorm(n, sd = 0.6)

df <- data.frame(education, skill, network, income)
```

## Running the search

The number of DAGs grows explosively with the number of variables, so we narrow
the space with background knowledge. Here we know only that `education` comes
first — nothing causes a person's education in this system — so we use
`disallowed` to forbid any edge pointing *into* `education`. The ordering among
`skill`, `network`, and `income` is left for the data to resolve. This is
exactly the kind of light domain knowledge a researcher can bring to a problem.

```{r discover}
disallowed <- list(
  c("skill",   "education"), 
  c("network", "education"), 
  c("income", "education")
)

result <- KaguModel$discover(df, disallowed = disallowed)
```

```{r print-result}
result
```

The search evaluated every DAG consistent with that constraint, but — thanks to
the Markov factorisation — only had to fit each unique `node ~ parents` model
once (far fewer fits than DAGs).

## The posterior over DAGs

`$summary()` ranks the structures by posterior probability. Each row is a whole
DAG; `posterior_prob` is $P(G \mid X)$.

```{r summary}
result$summary()
```

The histogram makes the shape of the posterior clear. With the true
data-generating DAG highlighted, we can see how much of the probability mass it
captures.

```{r plot}
result$plot(true_dag = true_dag)
```

## Edge probabilities

Often we care less about the single best graph than about *specific* causal
questions — does this edge exist, and in which direction? `$edge_probabilities()`
marginalises over the whole posterior to give the probability of each directed
edge, $P(\text{from} \to \text{to} \mid X) = \sum_{G \ni \text{edge}} P(G \mid X)$.

```{r edges}
result$edge_probabilities()
```

The four true edges (`education→skill`, `education→network`, `skill→income`,
`network→income`) should carry almost all the probability, while spurious edges
(a direct `education→income`, or `skill→network`) carry very little.

## Effects under structural uncertainty

When the structure is uncertain, committing to a single DAG — even the most
probable one — throws away that uncertainty. Instead, query the effect directly
on the `DiscoveryResult`: it averages over the *entire* posterior of DAGs,
weighting each by $P(G \mid X)$,
$$
P(Q \mid X) = \sum_G P(Q \mid X, G)\, P(G \mid X).
$$
The call signature and the returned object are exactly those of the single-DAG
`model$effects()`, so everything downstream — summaries, plots — works the same.

```{r effect}
result$effects("education", "income")$summary()
```

This estimate carries both the parameter uncertainty within each model and the
structural uncertainty between models. DAGs in which `education` has no causal
path to `income` simply contribute a zero effect, exactly as they should.

## The limits of observational data

Structure discovery is powerful, but it cannot manufacture information the data
do not contain — and a good method should tell you so. The sharpest example is
just two variables. Suppose `a` causes `b`:

```{r two-node}
# set.seed(42)
n <- 20000
a <- rnorm(n)
b <- 0.9 * a + rnorm(n, sd = 0.5)

KaguModel$discover(data.frame(a = a, b = b))$summary()
```

The search decisively rejects independence — `a` and `b` are plainly related —
but it splits almost exactly evenly between `a → b` and `b → a`. And it is
*right* to: the two graphs imply the identical observational distribution (they
are **Markov equivalent**), so no quantity of observational data can distinguish
them. This near-tie is not indecision that more data would resolve; collect a
million rows and it stays a tie. On any single dataset the split even wobbles
away from 50/50 by chance — rerun with fresh data and the coin lands the other
way — because there is genuinely no signal to extract about direction.

The same holds for any Markov-equivalent set, such as the chains `x → y → z`,
`x ← y ← z`, and the fork `x ← y → z`, which all imply `x ⊥ z | y`. Breaking the
tie requires stepping outside passive observation: a temporal ordering or other
domain knowledge (as in the worked example above), or an **intervention** —
randomising `a` and seeing whether `b` responds. Reporting the tie honestly,
rather than picking a winner, is exactly what makes the posterior-over-DAGs
framing trustworthy.

The uniform prior used here can, in future, be replaced by priors that favour
sparsity or encode softer structural beliefs — the posterior machinery is
unchanged.
