---
title: "Causal structure discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal structure discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4,
  message   = FALSE
)
library(kagu)
set.seed(42)
```

The other vignettes take the DAG as given. But the DAG is itself a *hypothesis*
about how a system works, and often it is exactly what we want to learn. Kagu
can treat structure discovery as Bayesian inference over graphs: fit a GCM to
every candidate DAG, and turn the marginal likelihoods into a posterior
distribution over structures,

$$
P(G \mid X) \;\propto\; P(X \mid G)\, P(G).
$$

Here $P(X \mid G)$ is the marginal likelihood of the data under graph $G$
(computed in closed form from the per-node Gaussian-process fits) and $P(G)$ is a prior over
structures (currently uniform). Rather than returning a single "best" graph,
this returns a full probability distribution — an honest representation of what
the data can and cannot tell us about causal structure.

## A worked example

We simulate a four-variable system with a known structure. `education` is a root
cause of both `skill` and `network`; `income` is driven by `skill` and
`network`. There is **no** direct `education → income` edge, and `skill` and
`network` are not directly linked.

```{r truth, fig.height = 3.5}
true_dag <- list(
  education = c(),
  skill     = "education",
  network   = "education",
  income    = c("skill", "network")
)

kagu_plot_dag(true_dag, node_pos = list(
  education = c(0,  0),
  skill     = c(1, -1),
  network   = c(1,  1),
  income    = c(2,  0)
))
```

```{r simulate}
n         <- 50
education <- rnorm(n)
skill     <- 0.8 * education + rnorm(n, sd = 0.6)
network   <- 0.6 * education + rnorm(n, sd = 0.6)
income    <- 1.0 * skill + 0.7 * network + rnorm(n, sd = 0.6)

df <- data.frame(education, skill, network, income)
```

## Running the search

The number of DAGs grows explosively with the number of variables, so we narrow
the space with background knowledge. Here we know only that `education` comes
first — nothing causes a person's education in this system — so we use
`disallowed` to forbid any edge pointing *into* `education`. The ordering among
`skill`, `network`, and `income` is left for the data to resolve. This is
exactly the kind of light domain knowledge a researcher can bring to a problem.

```{r discover}
disallowed <- list(
  c("skill",   "education"), 
  c("network", "education"), 
  c("income", "education")
)

result <- KaguModel$discover(df, disallowed = disallowed)
```

```{r print-result}
result
```

The search evaluated every DAG consistent with that constraint, but — thanks to
the Markov factorisation — only had to fit each unique `node ~ parents` model
once (far fewer fits than DAGs).

### Tighter constraints: `required` edges

For larger systems, the number of candidate DAGs can easily exceed Kagu's safety limits (capped at $2^{24}$ subsets, roughly 5 unconstrained nodes). If you have stronger structural beliefs—such as an edge that absolutely *must* exist—you can pass `required`. This shrinks the search space dramatically by removing those edges from the combinatorics.

```r
# Suppose we know education -> skill and education -> network are certain
result_req <- KaguModel$discover(
  df, 
  disallowed = disallowed, 
  required = list(c("education", "skill"), c("education", "network"))
)
```

### Explicit DAG comparison: `dags`

Sometimes you don't want to search the whole graph space. If you have a few specific, theoretically motivated hypotheses, you can pass them directly via `dags`. Kagu will skip the combinatorial search entirely and just compute the posterior probabilities over your explicit models.

```r
dag_mediation <- list(education = c(), skill = "education", income = "skill")
dag_direct    <- list(education = c(), skill = c(), income = "education")

result_explicit <- KaguModel$discover(df, dags = list(dag_mediation, dag_direct))
```

## The posterior over DAGs

`$summary()` ranks the structures by posterior probability. Each row is a whole
DAG, referenced by a short `id` (here the enumerated graphs get ids `"A"`, `"B"`,
`"C"`, …; had we passed a *named* list of DAGs, those names would be used
instead). `posterior_prob` is $P(G \mid X)$.

```{r summary}
result$summary()
```

To see what a given structure actually is, look it up by its id with
`result$get_dag(id)`, or plot it directly:

```{r inspect-dag, fig.height = 3.5}
top_id <- result$summary()$id[1]
result$plot_dag(top_id)
```

No single DAG runs away with the posterior — the top few structures each hold
only around a sixth of the mass. That is not a defect: with a handful of
variables and this much noise, the data simply cannot single out one graph. What
it *can* do is concentrate the mass onto a small cluster of closely-related
structures that all agree on the important causal edges, and rule the rest out.
We can't always name the one true DAG, but we can zoom in on the handful that
matter.

The histogram makes the shape of the posterior clear. With the true
data-generating DAG highlighted, we can see how much of the probability mass it
captures.

```{r plot}
result$plot(true_dag = true_dag)
```

## Edge probabilities

Often we care less about the single best graph than about *specific* causal
questions — does this edge exist, and in which direction? `$edge_probabilities()`
marginalises over the whole posterior to give the probability of each directed
edge, $P(\text{from} \to \text{to} \mid X) = \sum_{G \ni \text{edge}} P(G \mid X)$.

```{r edges}
result$edge_probabilities()
```

This is where the posterior becomes genuinely useful. The four true edges
(`education→skill`, `education→network`, `skill→income`, `network→income`) each
carry almost all of the probability — the data is confident they exist and point
the way they do — and the clearest edge *reversals* are correctly assigned low
probability. What lingers is a redundant direct `education→income` edge: even
though education's effect on income is *fully mediated* by skill and network,
observational data at this sample size cannot completely exclude a direct path,
so that edge keeps a moderate probability rather than collapsing to zero. Kagu
reports the residual ambiguity honestly instead of pretending the edge is ruled
out.

## Effects under structural uncertainty

When the structure is uncertain, committing to a single DAG — even the most
probable one — throws away that uncertainty. Instead, query the effect directly
on the `DiscoveryResult`: it averages over the *entire* posterior of DAGs,
weighting each by $P(G \mid X)$,
$$
P(Q \mid X) = \sum_G P(Q \mid X, G)\, P(G \mid X).
$$
The call signature and the returned object are exactly those of the single-DAG
`model$effects()`, so everything downstream — summaries, plots — works the same.

```{r effect}
result$effects("education", "income")$summary()
```

This estimate carries both the parameter uncertainty within each model and the
structural uncertainty between models. DAGs in which `education` has no causal
path to `income` simply contribute a zero effect, exactly as they should.

## The limits of observational data

Structure discovery is powerful, but it cannot manufacture information the data
do not contain — and a good method should tell you so. A classic example is a 
three-variable system with linear relationships. Suppose `x` causes `y`, which in turn causes `z`:

```{r three-node}
n <- 500
x <- rnorm(n)
y <- 0.8 * x + rnorm(n, sd = 0.5)
z <- 0.8 * y + rnorm(n, sd = 0.5)

res_linear <- KaguModel$discover(data.frame(x, y, z))
head(res_linear$summary())
```

The search decisively rejects the independent graph and any structure that
violates the data's conditional independencies, but it does **not** collapse onto
a single winner. Instead the mass is shared across a cluster of closely-related
graphs: the reverse chain (`y → x; z → y`), the fork (`y → x; y → z`), and the
true forward chain (`x → y; y → z`) — plus versions of each carrying one
redundant extra edge that the data cannot rule out.

And it is *right* to spread out. In a linear system, the three chains imply the
identical observational distribution (they are **Markov equivalent**). They all
encode that `x` and `z` are independent given `y`, and no quantity of purely
observational linear data can distinguish them — nor can it exclude a weak extra
edge that adds nothing to the fit. Reporting this uncertainty honestly, rather
than picking an arbitrary winner, is exactly what makes the posterior-over-DAGs
framing trustworthy.

(Note that strict Markov equivalence often relies on linear-Gaussian assumptions. 
If the underlying relationships were non-linear, Kagu's Gaussian-process mechanisms 
could often uniquely identify the exact true structure from observational data alone!)

Breaking the tie requires stepping outside passive linear observation: non-linear relationships, 
temporal ordering or domain knowledge (as in the worked example above), or an **intervention** — 
randomising `x` and seeing whether `y` and `z` respond.

The uniform prior used here can, in future, be replaced by priors that favour
sparsity or encode softer structural beliefs — the posterior machinery is
unchanged.
