---
title: "Modelling interactions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7, fig.height = 4.5,
  message   = FALSE,
  warning   = FALSE
)
library(kagu)
library(ggplot2)
set.seed(42)
```

In traditional linear modelling (`lm`), interactions between variables must be explicitly specified by the researcher (e.g., `y ~ a * b`). If you forget to include the interaction term, the model is forced to assume that the effect of `a` is strictly independent of `b`, leading to biased or incomplete estimates.

Kagu takes a different approach. Because Kagu models each node as a Gaussian Process over an eigen-basis of its parents, **interactions are learned automatically**. The GP naturally spans the multi-dimensional surface of its inputs, capturing non-linearities and interactions without any explicit formula from the user.

In this vignette, we'll look at how to recover and visualise a continuous-continuous interaction using Kagu's effect estimation.

## A system with a strong interaction

Let's simulate a system where $A$ and $B$ are independent causes of $Y$, but they interact multiplicatively. The true data-generating process is:
$$ Y = 1.5 A - 1.0 B + 2.0 (A \times B) + \epsilon $$

Because of the positive $2.0$ interaction term, the effect of $A$ on $Y$ heavily depends on the value of $B$. If $B$ is positive, $A$ has a strong positive effect. If $B$ is negative, $A$'s effect might even become negative!

```{r simulate}
n <- 500
a <- rnorm(n)
b <- rnorm(n)
y <- 1.5 * a - 1.0 * b + 2.0 * (a * b) + rnorm(n, sd = 0.5)

df <- data.frame(a, b, y)
```

We define our Kagu model as usual and fit it. Notice that we do *not* tell Kagu about the interaction.

```{r fit}
dag <- list(
  a = c(),
  b = c(),
  y = c("a", "b")
)

mod <- KaguModel$new(dag)
mod$fit(df)
```

## The marginal effect

If we ask Kagu for the effect of `a` on `y` without specifying anything else, it integrates out the uncertainty and natural distribution of `b` over the dataset.

```{r marginal-effect}
# The average marginal gradient of a -> y
mod$effects("a", "y")$summary()
```

The average effect is positive (around 1.5), which matches the main effect of `a` in our simulation. However, the standard deviation/HDI is quite wide. That's because the "true" effect of `a` is not a single number-it's highly variable depending on `b`.

## Conditioning to reveal the interaction

To probe the interaction, we can estimate the effect of `a` on `y` while **conditioning** `b` at specific values using the `conditions` argument.

Let's look at the gradient (the local slope) of `a -> y` when $B = -1$, $B = 0$, and $B = 1$:

```{r conditional-effects}
# Slope when b = -1
mod$effects("a", "y", conditions = list(b = -1))$summary()

# Slope when b = 0
mod$effects("a", "y", conditions = list(b = 0))$summary()

# Slope when b = 1
mod$effects("a", "y", conditions = list(b = 1))$summary()
```

The model successfully recovered the interaction! 
* At $B = -1$, the expected slope is $1.5 + 2.0(-1) = -0.5$. Kagu estimates $\approx -0.5$.
* At $B = 0$, the expected slope is $1.5 + 2.0(0) = 1.5$. Kagu estimates $\approx 1.5$.
* At $B = 1$, the expected slope is $1.5 + 2.0(1) = 3.5$. Kagu estimates $\approx 3.5$.

## Visualising the interaction with sweeps

We can visualise this interactive surface by requesting `sweep = TRUE`. Rather than computing the effects one at a time and combining them manually, we can pass a list of conditions directly to `$effects()`.

Kagu will automatically compute the sweep under each condition, bundle them together, and overlay them seamlessly when you call `$plot()`!

```{r plot-sweeps}
# Request full sweeps of a -> y at different levels of b
eff_multi <- mod$effects("a", "y", sweep = TRUE, conditions = list(
  list(b = -1),
  list(b = 0),
  list(b = 1)
))

# Plot the interaction seamlessly
eff_multi$plot() +
  scale_color_manual(values = c("b=-1" = "#ff7043", "b=0" = "#9e9e9e", "b=1" = "#26a69a")) +
  scale_fill_manual(values = c("b=-1" = "#ff7043", "b=0" = "#9e9e9e", "b=1" = "#26a69a")) +
  ggtitle("Continuous-Continuous Interaction: a -> y conditioned on b")
```

The diverging slopes clearly show the interaction effect.

By treating causal mechanisms as flexible Gaussian Processes, Kagu frees you from having to guess or pre-specify the exact functional form of your causal relationships. Interactions are automatically detected, naturally accommodated, and easily visualised using the `conditions` argument!
