---
title: "Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
library(kagu)
set.seed(42)
```

This quickstart fits a small causal model end to end: define the graph, fit it,
and read off causal effects with full posterior uncertainty.

## Define the causal structure

A Kagu model is a DAG - a named list mapping each node to its parent nodes.
Here `income` is caused by both `education` and `experience`, which have no
parents of their own (they are *root* nodes, written as `c()`).

```{r dag, fig.height = 3}
dag <- list(
  education = c(),
  experience = c(),
  income = c("education", "experience")
)

model <- KaguModel$new(dag = dag)

# node_pos gives each node a c(row, col) grid coordinate (row increases
# downward, col rightward) so the layout renders predictably.
model$plot_dag(node_pos = list(
  education  = c(0, -1),
  experience = c(0,  1),
  income     = c(1,  0)
))
```

The arrows encode our causal assumptions: an arrow `A → B` means "A is a direct
cause of B". Everything Kagu computes is relative to this structure, so it pays
to get it right.

## Simulate data and fit

We generate data in which each extra year of education adds £500 to income and
each year of experience adds £200, then fit the model. Each node's conditional
distribution is modelled as a Gaussian process of its parents, so the fit adapts
to whatever shape the relationship takes.

```{r data}
N          <- 500
education  <- rnorm(N, 14, 2)
experience <- rnorm(N, 10, 5)
income     <- 3000 + 500 * education + 200 * experience + rnorm(N, sd = 2000)

data <- data.frame(
  education  = education,
  experience = experience,
  income     = income
)
```

```{r fit, results = "hide", message = FALSE}
model$fit(data)
```

## Causal effects

The default `effects()` call returns the **local causal effect at the mean** -
the instantaneous slope of `target` with respect to `source`, holding the
intervention central. For a linear mechanism this is directly comparable to a
regression coefficient, and here it recovers the data-generating value of ≈ 500.

```{r unit-effect}
effect <- model$effects("education", "income")
effect$summary()
```

The summary is a tidy tibble. Reading the columns:

- `from` / `to` - the two intervention values being contrasted.
- `mean` - the posterior mean effect (our point estimate).
- `sd` - the posterior standard deviation (how uncertain that estimate is).
- `hdi_lower` / `hdi_upper` - the bounds of the 90% **highest density interval**
  (HDI): the narrowest interval containing 90% of the posterior probability.
  It is the Bayesian analogue of a confidence interval, but with the
  interpretation people usually *want*: there is a 90% probability the effect
  lies inside it, given the model and data. When the HDI excludes 0, the effect
  is credibly non-zero in that direction.

### Standardised effect

`std_units = TRUE` reports the effect of a **one standard-deviation** increase
in the cause, centred at its mean. This puts effects of variables measured on
different scales onto a common footing for comparison.

```{r sd-effect}
effect_sd <- model$effects("education", "income", std_units = TRUE)
effect_sd$summary()
```

### Explicit contrast

Pass `values = c(from, to)` to ask a concrete counterfactual question. Here:
how much more would someone with 16 years of education earn than someone with
12, all else flowing through the causal model?

```{r contrast}
effect_contrast <- model$effects("education", "income", values = c(12, 16))
effect_contrast$summary()
```

The mean is roughly four times the unit effect, as expected for a four-year gap
under a linear mechanism.

### Visualising the posterior

`plot()` shows the full posterior distribution of the effect - the point is the
mean and the bar beneath is the HDI. The spread *is* the uncertainty.

```{r plot-effect}
effect$plot()
```

## Dose-response curve

A sweep traces the expected outcome across a grid of intervention values,
`E[income | do(education = x)]`, with an HDI ribbon. This is the interventional
dose-response curve: what we expect income to be if we *set* education to each
value.

```{r sweep}
sweep <- model$effects("education", "income", sweep = TRUE)
sweep$plot()
```

## Model summary

`summary()` returns a table across every node. For each parent it reports the
**direct local effect** - the gradient of the node's fitted function with respect
to that parent, evaluated at the parents' means. For a linear relationship this
is exactly the slope, so here `education` and `experience` recover ≈ 500 and
≈ 200. Each node also gets a `sigma (noise)` row for its residual scale, and
everything carries a posterior mean, sd, and HDI.

```{r summary}
model$summary()
```

## Diagnostics

`diagnostics()` reports the posterior draw count and residual noise per node.
(The Gaussian-process posterior is available in closed form as a single sample,
so the chain-based r-hat / ESS diagnostics of MCMC do not apply.)

```{r diagnostics}
model$diagnostics("income")
```

## Save and load

A fitted model (optionally with its data) round-trips to disk, so you can refit
once and query effects later.

```{r save-load}
tmp <- tempfile(fileext = ".rds")
model$save(tmp)
loaded <- KaguModel$load(tmp)
loaded$effects("education", "income")$summary()
```
