---
title: "Markov Equivalence Stability across Sample Sizes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(kagu)
library(ggplot2)
library(dplyr)
```

## Objective

We want to investigate how stable the posterior probabilities are for two Markov-equivalent DAGs (`x -> y` vs `y -> x`) as the sample size $N$ increases. 

Because the underlying data-generating process is purely linear-Gaussian, the two models are theoretically indistinguishable. However, Kagu's Gaussian Process mechanism fits non-linear bases. We will generate a single dataset and incrementally "unveil" more data to see if the probability stabilises, converges, or remains erratic.

## Configuration

```{r config}
# Parameters
set.seed(42)
max_n    <- 300
start_n  <- 10
step_n   <- 5

beta     <- 0.8
noise_sd <- 0.5
```

## Data Generation

We generate the full dataset once so that we are strictly adding new information at each step, rather than drawing entirely new datasets.

```{r simulate}
# Generate the full dataset
x <- rnorm(max_n)
y <- beta * x + rnorm(max_n, sd = noise_sd)
df_full <- data.frame(x, y)

# Define the two candidate DAGs
dag_xy <- list(x = c(), y = "x")
dag_yx <- list(x = "y", y = c())
candidates <- list(dag_xy, dag_yx)
```

## Incremental Unveiling

We loop over the sample sizes, subsetting the data from $1$ to $N$, and compute the posterior probabilities.

```{r run-discovery}
# Sequence of sample sizes
ns <- seq(start_n, max_n, by = step_n)

# Suppress Kagu output for the loop
results <- withr::with_options(list(kagu.quiet = TRUE), {
  do.call(rbind, lapply(ns, function(n) {
    df_subset <- df_full[1:n, ]
    
    # Run discovery restricted to just the two DAGs
    res <- kagu_discover(df_subset, dags = candidates)
    
    # Extract the probabilities
    probs <- res$probabilities()
    
    data.frame(
      N = n,
      prob_xy = probs$posterior_prob[probs$edges == "x→y"],
      prob_yx = probs$posterior_prob[probs$edges == "y→x"]
    )
  }))
})
```

## Results and Visualisation

Let's plot the posterior probability of `x -> y` (and implicitly `y -> x`) as the sample size grows.

```{r plot-stability, fig.width=8, fig.height=5}
# Convert to long format for ggplot
plot_df <- tidyr::pivot_longer(
  results, 
  cols = c("prob_xy", "prob_yx"),
  names_to = "Direction",
  values_to = "Probability"
) %>%
  mutate(Direction = ifelse(Direction == "prob_xy", "x -> y", "y -> x"))

ggplot(plot_df, aes(x = N, y = Probability, color = Direction)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_color_manual(values = c("x -> y" = "#26a69a", "y -> x" = "#ff7043")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Stability of Markov Equivalent DAGs over Sample Size",
    subtitle = paste0("Linear data generating process (x -> y). Maximum N = ", max_n),
    x = "Sample Size (N)",
    y = "Posterior Probability"
  ) +
  theme(legend.position = "top")
```

### Interpretation

Because the data is perfectly linear, the two models are Markov equivalent. If they were evaluated using strictly linear models, their marginal likelihoods would be mathematically identical, yielding exactly a 50/50 split. 

However, because Kagu's GP mechanism searches for non-linear structures, it behaves like a highly sensitive magnifying glass over the random noise in the finite sample. As we unveil more data, the random empirical skewness and "wiggles" in the data continually shift, occasionally making one direction temporarily look smoother or more natural to the GP than the other. 

Therefore, even as $N$ grows, the probabilities do not confidently converge to 100% for `x -> y`, nor do they perfectly flatline at 50%. Instead, they **erratically wobble** between 0 and 1, faithfully reflecting the fundamental lack of identifiability (i.e. the true signal difference is exactly zero, so we are observing $O(1)$ noise dominating the likelihood ratio).